Models of VOR learning in MHC knockout mice

### 2013-01-26 Notes/VORmodel.tex
pooled model figs

We make the following assumptions:
\begin{itemize}
  \item There are $N$ identical synapses with $M$ internal functional states.
  \item The states of different synapses are independent of each other.
  \item The synapses that are eligible for plasticity are chosen randomly.
  \item The potentiating/depressing plasticity event timings are distributed as Poisson processes with rates $rf\potdep$, where $f\pot+f\dep=1$.
  \item Potentiation and depression are described by Markov processes with transition probabilities $\M\potdep$.
  \item The synaptic weights of the internal states are given by the column vector $\w$. This can only take two values that we can call $\pm1$, except for the pooled resource model that we will discuss in \sref{sec:pooledmodel} and \sref{sec:pooled}.
\end{itemize}

@@

The untrained case will be described by $f\pot=f\pot\norm$.
Gain-increase training  will be described by $f\pot=f\pot\inc<f\pot\norm$, and
gain-decrease training  will be described by $f\pot=f\pot\dec<f\pot\norm$. Note that the forgetting matrix \eqref{eq:evolve} and the equilibrium distribution \eqref{eq:eqprob} depends on $f\pot$, which we will indicate with subscripts.

@@

If they are equal, this would change the mean synaptic weight in equilibrium.
This seems like it would affect the ability of the network to perform its function, and one might expect adaptation to the environment to produce an equilibrium state that has the same performance.
In any case, if the synaptic weights are different, the electrical activity will be different, and there will be no reason to expect the same statistics for potentiation or depression.

On the other hand, one could imagine adjusting $f\pot\wt$ and $f\pot\ko$ so that $\eq\wt\w = \eq\ko\w$.
But, now that the synaptic weights are the same, the electrical activity will be the same, and there will be no reason to expect different statistics for potentiation or depression.
If we did this for the multistate or two-state models, we'd remove all differences between the wild-type and \KO\ mutant.

@@

We will look at three different models, the cascade model (see \cite{Fusi2005cascade} and \autoref{fig:models}\ref{fig:cascade}), the multistate model (see \cite{amit1994learning,Fusi2007multistate} and \autoref{fig:models}\ref{fig:multistate}) and the two-state model (which can be thought of as a special case of the previous two, see \autoref{fig:models}\ref{fig:binary}).

@@

For the multistate and two state models, we will use the same value for the transition probabilities, $q$ for potentiation and depression in the wild-type as well as potentiation in the \KO\ mutant.
We will use a larger value for $q$ for depression in the \KO\ mutant.

In each case, we set $f\pot\norm=\half$, $f\pot\inc=f\pot\norm+\Delta f$ and $f\pot\dec=f\pot\norm-\Delta f$. The values of all these parameters are listed in \autoref{tab:params}.

@@

\begin{table}
 \begin{center}
  \begin{tabular}{|l|c|c|c|c|c|c|}
    \hline
    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
    Model & \# states & pot,WT dep & \KO\ dep & $\Delta f$ & $r\ttrain$ & $r\tpre$ \\
    \hline
    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 & 50 & 50 \\
    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 & 50 & 150 \\
    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.1 & 50 & 50 \\
    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.4 & 20 & 20 \\
    Two-state  & 2  & $q=0.4$  & $q=0.8$  & -0.1 & 5 & 5 \\
    \hline
  \end{tabular}
 \end{center}
  \caption{Parameters used in simulations.} \label{tab:params}
\end{table}

@@

@@

The results of simulations of the multistate model can be seen in \autoref{fig:binary_res}.
However, this model can be solved exactly:
%
\begin{multline}\label{eq:binarysol}
  \eq = \frac{(f\dep q\dep, f\pot q\pot)}{\lambda},
  \qquad
  \pr(t) = \eq + (\pr(t)-\eq)\e^{-\lambda rt},\\
  \qquad \text{where} \quad
  \lambda = f\pot q\pot + f\dep q\dep.
\end{multline}
%
But it is easier to just substitute $M=2$ into the formulae in \sref{sec:multistate}.
In this case, the initial rate of change encapsulates the whole solution, as there is only a single exponential decay.

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:binWTnopre}
  \Phi = -\Delta f q,
\end{equation}
%
where, again, it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:binWTpre}
\begin{aligned}
  \Phi &= -2\Delta f q.
\end{aligned}
\end{equation}
%
So, we see that pre-training will always speed up learning, unlike what is seen in the experiment.
This can be seen in \autoref{fig:binary_res}\ref{fig:binary_learn},\ref{fig:binary_learnS}.

@@

>>>>

We make the following assumptions:
\begin{itemize}
  \item There are $N$ identical synapses with $M$ internal functional states.
  \item The states of different synapses are independent of each other.
  \item The synapses that are eligible for plasticity are chosen randomly.
  \item The potentiating/depressing plasticity event timings are distributed as Poisson processes with rates $rf\potdep$, where $f\pot+f\dep=1$.
      In other words, $r$ is the total rate of plasticity events per synapse and $f\potdep$ are the fraction of these events that are potentiating/depressing.
  \item Potentiation and depression are described by Markov processes with transition probabilities $\M\potdep$.
  \item The synaptic weights of the internal states are given by the column vector $\w$. This can only take two values that we can call $\pm1$, except for the pooled resource model that we will discuss in \autoref{sec:pooledmodel} and \autoref{sec:pooled}.
\end{itemize}

@@

The untrained case will be described by $f\pot=f\pot\norm$.
Gain-increase training  will be described by $f\pot=f\pot\inc<f\pot\norm$, and
gain-decrease training  will be described by $f\pot=f\pot\dec<f\pot\norm$. Note that the forgetting matrix \eqref{eq:evolve} and the equilibrium distribution \eqref{eq:eqprob} depend on $f\pot$, which we will indicate with subscripts.

@@

If they are equal, this would change the mean synaptic weight in equilibrium.
This seems like it could affect the ability of the network to perform its function, whereas, in reality, the wild-type and mutant have the same baseline performance.
In any case, if the synaptic weights are different, the electrical activity will be different, and there will be no reason to expect the same statistics for potentiation or depression.
In setting $f\pot\wt=f\pot\ko$, we are assuming that the rest of the brain compensates for the difference in synaptic weights somehow.

On the other hand, one could imagine adjusting $f\pot\wt$ and $f\pot\ko$ so that $\eq\wt\w = \eq\ko\w$.
But, now that the synaptic weights are the same, the electrical activity will be the same, and there will be no reason to expect different statistics for potentiation or depression.
If we did this for the multistate or two-state models, we'd remove all differences between the wild-type and \KO\ mutant, so we won't.

@@

We will look at three different models, the cascade model (see \cite{Fusi2005cascade} and \autoref{fig:models}\ref{fig:cascade}), the multistate model (see \cite{amit1994learning,Fusi2007multistate} and \autoref{fig:models}\ref{fig:multistate}) and the two-state model (which can be thought of as a special case of the previous two, see \autoref{fig:models}\ref{fig:binary}).
We will also look at a new, pooled resource model that we will define below in \autoref{sec:pooledmodel}.

@@

For the multistate and two state models, we will use the same value for the transition probabilities, $q$, for potentiation and depression in the wild-type as well as potentiation in the \KO\ mutant.
We will use a larger value for $q$ for depression in the \KO\ mutant.

In each case, we set $f\pot\norm=\half$, $f\pot\inc=f\pot\norm+\Delta f$ and $f\pot\dec=f\pot\norm-\Delta f$. 
We use the same values for wild-type and mutant and treat gain-increase and decrease symmetrically.
The values of all these parameters are listed in \autoref{tab:params}.

@@

\begin{table}
 \begin{center}
  \begin{tabular}{|l|c|c|c|c|c|c|}
    \hline
    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
    Model & \# states & pot,WT dep & \KO\ dep & $\Delta f$ & $r\ttrain$ & $r\tpre$ \\
    \hline
    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 & 50 & 50  \\
    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 & 50 & 150 \\
    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.1 & 50 & 50  \\
    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.4 & 20 & 20  \\
    Two-state  & 2  & $q=0.4$  & $q=0.8$  & -0.1 & 5  & 5   \\
    Pooled     & 10 & $q\in[0.3,0.4]$  & $q\in[0.6,0.8]$
                                          & -0.1 & 70 & 70 \\
    Pooled     & 10 & $q\in[0.1,0.4]$  & $q\in[0.2,0.8]$
                                          & -0.1 & 70 & 70 \\
    \hline
  \end{tabular}
 \end{center}
  \caption{Parameters used in simulations.} \label{tab:params}
\end{table}

@@

This model is parameterised by the range of values, $q\in[q\lmin,q\lmax]$, for potentiation and depression.
We will use the same values for potentiation and depression in the wild-type as well as potentiation in the \KO\ mutant.
We will use larger values for depression in the \KO\ mutant.

@@

The results of simulations of the two-state model can be seen in \autoref{fig:binary_res}.
However, this model can be solved exactly:
%
\begin{multline}\label{eq:binarysol}
  \eq = \frac{(f\dep q\dep, f\pot q\pot)}{\lambda},
  \qquad
  \pr(t) = \eq + (\pr(t)-\eq)\e^{-\lambda rt},\\
  \qquad \text{where} \quad
  \lambda = f\pot q\pot + f\dep q\dep.
\end{multline}
%
But it is easier to just substitute $M=2$ into the formulae in \autoref{sec:multistate}.
In this case, the initial rate of change encapsulates the whole solution, as there is only a single exponential decay.

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:binWTnopre}
  \Phi = -\Delta f q,
\end{equation}
%
where, again, it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:binWTpre}
\begin{aligned}
  \Phi &= -2\Delta f q.
\end{aligned}
\end{equation}
%
So, we see that pre-training will always speed up learning, unlike what is seen in the experiment.
This can be seen in \autoref{fig:binary_res}\ref{fig:binary_learnS}.

@@

\begin{figure}
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=7cm]{pooled_plenty_learn.eps}}\label{fig:pooled_plenty_learn}
  \item\aligntop{\includegraphics[width=7cm]{pooled_plenty_learnS.eps}}\label{fig:pooled_plenty_learnS}
  \item\label{fig:pooled_plenty_eq}\begin{myenumi}
                    \item\aligntop{\includegraphics[width=7cm]{pooled_plenty_eq_WT.eps}}\label{fig:pooled_plenty_eq_WT}
                    \item\aligntop{\includegraphics[width=7cm]{pooled_plenty_eq_KO.eps}}\label{fig:pooled_plenty_eq_KO}
                  \end{myenumi}
  \item\label{fig:pooled_plenty_pr}\begin{myenumi}
                    \item\aligntop{\includegraphics[width=3cm]{pooled_plenty_pr_WT_nopre.eps}}\label{fig:pooled_plenty_pr_WT_nopre}
                    \item\aligntop{\includegraphics[width=3cm]{pooled_plenty_pr_WT_pre.eps}}\label{fig:pooled_plenty_pr_WT_pre}
                    \item\aligntop{\includegraphics[width=3cm]{pooled_plenty_pr_KO_nopre.eps}}\label{fig:pooled_plenty_pr_KO_nopre}
                    \item\aligntop{\includegraphics[width=3cm]{pooled_plenty_pr_KO_pre.eps}}\label{fig:pooled_plenty_pr_KO_pre}
                  \end{myenumi}
 \end{myenuma}
 \end{center}
  \caption{Simulation results for pooled resource model with light resource depletion ($q\lmin/q\lmax=0.75$).
  Parameters can be found in \autoref{tab:params}.
  (\ref{fig:pooled_plenty_learn}) Learning curves for wild-type and \KO\ mutant with and without pre-training.
  (\ref{fig:pooled_plenty_learnS}) Learning curves restricted to gain-increase training.
  (\ref{fig:pooled_plenty_eq}) Equilibrium distributions without training or with gain-increase/decrease training for (\ref{fig:pooled_plenty_eq_WT}) wild-type and (\ref{fig:pooled_plenty_eq_KO}) \KO\ mutant.
  (\ref{fig:pooled_plenty_pr}) Evolution of probability distributions for (\ref{fig:pooled_plenty_pr_WT_nopre}\ref{fig:pooled_plenty_pr_WT_pre}) wild-type and  (\ref{fig:pooled_plenty_pr_KO_nopre}\ref{fig:pooled_plenty_pr_KO_pre}) \KO\ mutant without (\ref{fig:pooled_plenty_pr_WT_nopre},\ref{fig:pooled_plenty_pr_KO_nopre}) and with (\ref{fig:pooled_plenty_pr_WT_pre},\ref{fig:pooled_plenty_pr_KO_pre}) pre-training. } \label{fig:pooled_plenty}
\end{figure}

\begin{figure}
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=7cm]{pooled_scarce_learn.eps}}\label{fig:pooled_scarce_learn}
  \item\aligntop{\includegraphics[width=7cm]{pooled_scarce_learnS.eps}}\label{fig:pooled_scarce_learnS}
  \item\label{fig:pooled_scarce_eq}\begin{myenumi}
                    \item\aligntop{\includegraphics[width=7cm]{pooled_scarce_eq_WT.eps}}\label{fig:pooled_scarce_eq_WT}
                    \item\aligntop{\includegraphics[width=7cm]{pooled_scarce_eq_KO.eps}}\label{fig:pooled_scarce_eq_KO}
                  \end{myenumi}
  \item\label{fig:pooled_scarce_pr}\begin{myenumi}
                    \item\aligntop{\includegraphics[width=3cm]{pooled_scarce_pr_WT_nopre.eps}}\label{fig:pooled_scarce_pr_WT_nopre}
                    \item\aligntop{\includegraphics[width=3cm]{pooled_scarce_pr_WT_pre.eps}}\label{fig:pooled_scarce_pr_WT_pre}
                    \item\aligntop{\includegraphics[width=3cm]{pooled_scarce_pr_KO_nopre.eps}}\label{fig:pooled_scarce_pr_KO_nopre}
                    \item\aligntop{\includegraphics[width=3cm]{pooled_scarce_pr_KO_pre.eps}}\label{fig:pooled_scarce_pr_KO_pre}
                  \end{myenumi}
 \end{myenuma}
 \end{center}
  \caption{Simulation results for pooled resource model with heavy resource depletion ($q\lmin/q\lmax=0.25$).
  Parameters can be found in \autoref{tab:params}.
  (\ref{fig:pooled_scarce_learn}) Learning curves for wild-type and \KO\ mutant with and without pre-training.
  (\ref{fig:pooled_scarce_learnS}) Learning curves restricted to gain-increase training.
  (\ref{fig:pooled_scarce_eq}) Equilibrium distributions without training or with gain-increase/decrease training for (\ref{fig:pooled_scarce_eq_WT}) wild-type and (\ref{fig:pooled_scarce_eq_KO}) \KO\ mutant.
  (\ref{fig:pooled_scarce_pr}) Evolution of probability distributions for (\ref{fig:pooled_scarce_pr_WT_nopre}\ref{fig:pooled_scarce_pr_WT_pre}) wild-type and  (\ref{fig:pooled_scarce_pr_KO_nopre}\ref{fig:pooled_scarce_pr_KO_pre}) \KO\ mutant without (\ref{fig:pooled_scarce_pr_WT_nopre},\ref{fig:pooled_scarce_pr_KO_nopre}) and with (\ref{fig:pooled_scarce_pr_WT_pre},\ref{fig:pooled_scarce_pr_KO_pre}) pre-training. } \label{fig:pooled_scarce}
\end{figure}

The results of simulations of the pooled resource model can be seen in \autoref{fig:pooled_plenty} and \autoref{fig:pooled_scarce}.

If we compare gain-increase learning in the mutant to wild-type, there are two effects: the increased transition rates speed up learning, but the equilibrium distribution is shifted to the depressed side (compare \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_eq}: \ref{fig:pooled_plenty_eq_WT} and \ref{fig:pooled_plenty_eq_KO} or \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_eq}: \ref{fig:pooled_scarce_eq_WT} and \ref{fig:pooled_scarce_eq_KO}), where there are fewer synapses available for depression and resources are depleted.
When resource depletion is less severe, the first effect dominates and the mutant learns faster than wild-type (see \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_learn},\ref{fig:pooled_plenty_learnS}).
However, when resource depletion is more severe, the second effect dominates and the mutant learns slower than wild-type (see \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_learn},\ref{fig:pooled_scarce_learnS}), which matches what is seen in the experiment.

Gain-decrease pre-training lessens the second effect, and results in the mutant learning faster than wild-type (see \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_learn},\ref{fig:pooled_plenty_learnS} or \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_learn},\ref{fig:pooled_scarce_learnS}), as seen in the experiment.

Gain-decrease pre-training will shift the distribution to the potentiated site, where there are more synapses available for depression and resources are more plentiful, for both the mutant and wild type.
This means that the pre-trained animal will learn faster then the untrained one for both mutant and wild-type (see \autoref{fig:pooled_plenty}\ref{fig:pooled_plenty_learnS} or \autoref{fig:pooled_scarce}\ref{fig:pooled_scarce_learnS}).
This differs from what is seen experimentally, where the pre-trained wild-type learns slower than the untrained one.
